---
title: "short-term adaptation and robustness"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---

The following script was used to calculate the percentage of improvement of the maximum specific growth rate after re-inoculation of yeast cell cultures from a 96- well plate with different conditions in exactly the same setup (after 48h). Correlations between different phenotypes and their robustness and percentage of improvement of µmax were also analysed. 

Load libraries
```{r}
library(uwot)
library(tidyverse)
library(ggpubr)
library(GGally)
```

Load data
```{r, warnings=F}
list_csv_files <- list.files(path = "~/processed_data")
df <- readr::read_csv(list_csv_files) %>%
  filter(Yx <= 0.61) %>% #theoretical maximum is 0.51 on glucose aerobic +- 0.1
  filter(Yp <= 0.61) %>% #theoretical maximum is 0.51 on glucose aerobic +- 0.1
  filter(CDW <= 39.65) %>% #assuming maximum Yx/s of 0.61 max biomass from 65 g/L glucose media is x = Yx/s(Si-S)=39.65 g/L 
  select(`condition`, `replicate`, `group`, `µmax2`, `µmax`,`strain`) %>%
  dplyr::rename("umax"="µmax")%>%
  dplyr::rename("umax2"="µmax2")%>%
  mutate(Pi=(umax2-umax)/umax*100)

#setting Inf values to max increment 
max <- max(df$Pi[is.finite(df$Pi)]) #==1321.83
df[c('Pi')][sapply(df[c('Pi')], is.infinite)] <- max 

```
# plot 

Visualizing percentage of improvement (all datapoints)

```{r}
ggplot(df, aes(x=strain, y=Pi)) +
  geom_violin(alpha=0.5, color="gray")+
  geom_point(aes(colour = factor(group), alpha=0.9))+
  # geom_boxplot()+
  # facet_wrap(~strain,  ncol=1, strip.position = "left", scales ="free_y")+
  scale_color_manual(values = c("acid" = "#00AFBB", "alcohol" = "#3581d8", "hexose" = "#fba465", "pentose" = "#FC4E07", "salt" = "grey", "aldehyde" = "#f2c85b"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
Creating a dataset with mean across perturbations of the %P increase 
```{r}
df_Ps <- df %>%
  group_by(strain, replicate) %>%
 dplyr::summarise(meanP = mean(Pi, na.rm=TRUE))
df_Ps <- df_Ps %>%
  group_by(strain)%>%
  mutate(mean=mean(meanP))%>%
  mutate(sdP=sd(meanP)) %>%
  filter(replicate==1) %>%
  select(-`replicate`)
head(df_Ps)
```
Creating a dataset with mean across strains of the %P increase
```{r}
df_Pp <- df %>%
  group_by(condition, replicate) %>%
  dplyr::summarise(meanP = mean(Pi, na.rm=TRUE))
df_Pp <- df_Pp %>%
  group_by(condition)%>%
  mutate(mean=mean(meanP))%>%
  mutate(sdP=sd(meanP)) %>%
  filter(replicate==1) %>%
  select(-`replicate`)
head(df_Pp)
```
Creating a dataset with mean across strains an conditions of the %P increase and plotting to visualize differences 

```{r}
df_Pcs <- df %>%
    group_by(group, strain)%>%
    dplyr::summarise(mean_Pi = mean(Pi, na.rm=TRUE)) 
head(df_Pcs)

ggplot(df_Pcs, aes(y = mean_Pi, x = strain, fill=group)) + 
  geom_bar(position="dodge",stat="identity") +
  theme_light()+
  scale_fill_manual(values = c("acid" = "#00AFBB", "alcohol" = "#3581d8", "hexose" = "#fba465", "pentose" = "#FC4E07", "salt" = "grey", "aldehyde" = "#f2c85b"))+
  # facet_grid(rows = "group", scales="free")+
  theme(
    text = element_text(size = 20),
    axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1),
    axis.title = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 20, color = "black", face = "bold"),
    strip.text.y = element_text(size = 20, color = "black", face = "bold"))

#ggsave("Pimprov_strains_groups_facet.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")

```
Plot data on perturbations and strains
```{r}
df_Ps <- df_Ps %>%
  mutate(strain = factor(
    strain,
    levels = c(
      "ETRED",
      "PE2",
      "Redstar",
      "Thermosacc",
      "CENPK7D",
      "S288C",
      "LBCM1001",
      "LBCM1003",
      "LBCM1008",
      "LBCM1013",
      "LBCM1014",
      "LBCM1017",
      "LBCM103",
      "LBCM1030",
      "LBCM1046",
      "LBCM1079",
      " LBCM109",
      "LBCM1095",
      "LBCM1099",
      "LBCM110",
      "LBCM1106",
      "LBCM37",
      "LBCM67",
      "LBCM97",
      ordered = T
    )
  )) 
df_Pcs <- df_Pcs %>%
  mutate(strain = factor(
    strain,
    levels = c(
      "ETRED",
      "PE2",
      "Redstar",
      "Thermosacc",
      "CENPK7D",
      "S288C",
      "LBCM1001",
      "LBCM1003",
      "LBCM1008",
      "LBCM1013",
      "LBCM1014",
      "LBCM1017",
      "LBCM103",
      "LBCM1030",
      "LBCM1046",
      "LBCM1079",
      " LBCM109",
      "LBCM1095",
      "LBCM1099",
      "LBCM110",
      "LBCM1106",
      "LBCM37",
      "LBCM67",
      "LBCM97",
      ordered = T
    )
  )) 

ggplot() +
  geom_bar(data=df_Ps, aes(x=strain, y=meanP,), stat="identity")+
  geom_errorbar(data=df_Ps, aes(x=strain, ymin=meanP-sdP, ymax=meanP+sdP), width=0.05, colour="black", alpha=0.9, size=0.5)+ 
  geom_point(data= df_Pcs, aes(x=strain, y=mean_Pi, colour = group, alpha=5), shape=20, size= 7)+
  scale_color_manual(values = c("acid" = "#00AFBB", "alcohol" = "#3581d8", "hexose" = "#fba465", "pentose" = "#FC4E07", "salt" = "grey", "aldehyde" = "#f2c85b"))+ 
  theme_light()+
  theme(
    text = element_text(size = 20),
    axis.text = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1),
    axis.title = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 20, color = "black", face = "bold"),
    strip.text.y = element_text(size = 20, color = "black", face = "bold"))

#ggsave("Pimprov_strains_condition.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")
  
ggplot(df_Pp, aes(x = condition, y = meanP)) + 
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(x=condition, ymin=meanP-sdP, ymax=meanP+sdP), width=0.05, colour="black", alpha=0.9, size=0.5)+ #error among triplicates
  theme_light()+
  theme(
    text = element_text(size = 20),
    axis.text = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1),
    axis.title = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 20, color = "black", face = "bold"),
    strip.text.y = element_text(size = 20, color = "black", face = "bold")) 

#ggsave("Pimprov_conditions.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")

```
# µmax correlations 
Check correlations between the maximum specific growth rate in the first and second cultivation

```{r}
#check correlation with spearman method
p <- ggscatter(df, x = 'umax', y = 'umax2', color = 'group', add = "reg.line", conf.int = FALSE, size = 5, palette=c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"))+
  stat_cor(aes(color = group), method = "spearman") 
p +
  font("xlab", size = 30)+
  font("ylab", size = 30)+
  font("xy.text", size = 30)

#ggsave("correlation_µmax_µmax2.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")
```

# correlations between robustness and adaptation 
```{r}
#uploading data with all phenotypic variables
df_a <- readr::read_csv(list_csv_files) %>%
  filter(Yx <= 0.61) %>% #theoretical maximum is 0.51 on glucose aerobic +- 0.1
  filter(Yp <= 0.61) %>% #theoretical maximum is 0.51 on glucose aerobic +- 0.1
  filter(CDW <= 39.65) %>% #assuming maximum Yx/s of 0.61 max biomass from 65 g/L glucose media is x = Yx/s(Si-S)=39.65 g/L 
  dplyr::rename("umax"="µmax")%>%
  dplyr::rename("umax2"="µmax2")%>%
  mutate(Pi=(umax2-umax)/umax*100)
#setting Inf values to max increment 
max(df_a$Pi[is.finite(df_a$Pi)]) #==1321.83
df_a[c('Pi')][sapply(df_a[c('Pi')], is.infinite)] <- 1321.83


#Calculating the mean and the standard deviation for each variable, strain and replicate and condition
df_P <- df_a %>%
  gather(key="phenotypes", value="value", c("umax", "lag", "Yx", "Yp", "CDW","Pi", "umax2")) %>%
  group_by(strain, phenotypes, condition) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) %>%
  mutate(stdev=sd(value,  na.rm=TRUE)) %>%
  filter(replicate == "1" ) %>%
  select(-c(group, value, r.c_µmax))

#calculating robustness 
df_mean <- df_a %>%
  gather(key="phenotypes", value="value", c("umax", "lag", "Yx", "Yp", "CDW")) %>%
  group_by(strain, phenotypes, replicate) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) %>%
  mutate(stdev=sd(value,  na.rm=TRUE)) %>%
  filter(condition == "Glu_20" ) %>%
  select(-c(condition, group, value))

df_R <- df_mean %>%
  group_by(phenotypes)%>%
  mutate(mean_v=mean(mean, na.rm=TRUE)) %>%
  mutate(robustness=-stdev^2/mean, na.rm=TRUE)%>%
  mutate(robustness_n=robustness/mean_v, na.rm=TRUE) %>%
  ungroup()%>%
  group_by(strain, phenotypes) %>%
  mutate(mean_R=mean(robustness_n, na.rm=TRUE)) %>%
  mutate(sem_R=sd(robustness_n)/sqrt(3),na.rm=TRUE) 
```

# more general correlations 
```{r}
#general correlation between adaptation and robusntess 
P_mean <- df_mean %>%
    dplyr::summarise(mean_Pi = mean(Pi, na.rm=TRUE))
P_mean <- P_mean %>%
  group_by(phenotypes, strain) %>%
  summarise_at(vars(mean_Pi), list(mean_Pi=mean)) 
R_mean <- df_R %>%
  dplyr::summarise(mean_R = mean(mean_R, na.rm=TRUE))

adp_R <- P_mean %>%
  merge.data.frame(R_mean, by=c("strain", "phenotypes")) %>%
  pivot_wider(names_from = phenotypes, values_from = mean_R) 

ggpairs(adp_R, columns = c(2:7),upper = list(continuous = wrap("cor", method = "spearman"))) 
# #ggsave("correlation_Pi_R.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")

#check percentage of improvement and robustness of specific phenotypes (change umax with the phenotype of interest)
ggscatter(adp_R, x = 'mean_Pi', y = 'umax',  add = "reg.line", conf.int = FALSE, size = 5)+
  stat_cor( method = "spearman") 
#ggsave("correlation_adap_Rumax.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")
```


