---
title: "Exploratory analysis of strain data"
author: "Ceclia Trivellin"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: kable
    toc: true
    toc_float: true
---
The following script is used to analyse the dataset containing phenotypic data (max.specific growth rate, lag phase, biomass and ethanol yield and cell dry weight) of 24 Saccharomyces cerevisiae strains cultivated in 29 conditions (triplicates). 

Load libraries
```{r, warning=F, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggsci)
library(RColorBrewer)
library(ggridges)
library(stats)
library(gghighlight)
library(ggbeeswarm)
library(uwot)
library(ggpubr)
library(GGally)
```

Load data
```{r}
df_strain_long <- readRDS("~/data/dataset.rds")
```

# Plotting strain phenotypes

Visualize data for each strain and phenotype
```{r}
ggplot(df_strain_long, aes(x=strain, y=value, order=strain_type)) +
  geom_violin(alpha=0.5, color="gray")+
  geom_point(aes(colour = factor(group), alpha=0.9))+
  # geom_boxplot()+
  facet_wrap(~phenotypes,  ncol=1, strip.position = "left", scales ="free_y")+
  scale_color_manual(values = c("acid" = "#00AFBB", "alcohol" = "#3581d8", "hexose" = "#fba465", "pentose" = "#FC4E07", "salt" = "grey", "aldehyde" = "#f2c85b"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

Visualize only the means for each group of conditions and for each strain 
```{r}
df_strain <-  df_strain_long %>% pivot_wider(names_from = phenotypes,values_from = value)

#calculating different means 
#datasets for plotting
ggdot_mean <- df_strain %>%
  gather(key="phenotypes", value="value", c("umax", "lag", "Yx", "Yp", "CDW")) %>%
  group_by(strain, phenotypes, group ) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) 
head(ggdot_mean)
ggdot_mean2 <- ggdot_mean %>%
  group_by(strain, phenotypes ) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) %>%
  filter(condition=="Glu_30", replicate==1)
head(ggdot_mean2)
#calculating means to compare strains, groups and conditions
mean_phenotype_group <- df_strain %>%
  gather(key="phenotypes", value="value", c("umax", "lag", "Yx", "Yp", "CDW")) %>%
  group_by(phenotypes, group ) %>%
 dplyr::summarise(mean_value = mean(value, na.rm=TRUE))
head(mean_phenotype_group)
phenotype_mean <- ggdot_mean %>%
  group_by(phenotypes ) %>%
  mutate(mean=mean(value,  na.rm=TRUE)) %>%
  filter(strain=="CENPK7D",condition=="Glu_30", replicate==1)
head(phenotype_mean)
condition_mean <- ggdot_mean %>%
  group_by(phenotypes, condition) %>%
  mutate(mean=mean(value,  na.rm=TRUE))%>%
  filter(replicate==1,  strain=="CENPK7D")
head(condition_mean)

p <- ggdotchart(ggdot_mean, x = "strain", y = "mean",
           group = "strain_type",
           color = "group",                                # Color by groups
           palette = c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"), # Custom color palette                           # Add segments from y = 0 to dots
           add.params = list(color = "lightgray", size = 2), # Change segment color and size
           dot.size = 2,                                 # Large dot size
           ggtheme = theme_light()                        # ggplot2 theme
)+
  geom_hline(yintercept = 0, linetype = 2, color = "lightgray") +
  facet_grid(vars(phenotypes), scales = "free_y")+
  theme(
    text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 20, color = "black", face = "bold"),
    strip.text.y = element_text(size = 20, color = "black", face = "bold")
  )

p2 <- p +
  geom_point(data = ggdot_mean2,aes(x=strain, y=mean), shape=95, size= 7)+
  facet_grid(vars(phenotypes), scales = "free_y")
p2

#ggsave("phenotype_mean.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")

```
Create metadata
```{r}
df_meta <- 
  df_strain_long %>% 
  select(Sample,strain,strain_type,condition,replicate,group) %>% 
  distinct()
```

# PCA
## Pre-processing

Let's begin by log-normalizing the data
```{r}
df_strain_long %>% 
  mutate(value_log = log10(value+1e-4)) %>% View()

pca_data <- 
  df_strain_long %>% 
  mutate(value_log = log10(value+1e-4)) %>% 
  select(-value) %>% 
  pivot_wider(names_from = phenotypes,values_from = value_log)

head(pca_data)
```

Make data suitable for prcomp function
```{r}
df_pca <- 
  pca_data %>% 
  select(Sample,CDW,Yx,Yp,umax,lag) %>% 
  as.data.frame()

rownames(df_pca) <- df_pca$Sample

df_pca <- 
  df_pca %>% 
  select(-Sample)
```

Run PCA
```{r}
pca1 <- prcomp(na.omit(df_pca),center = T,scale. = F)

pca1$x %>% head()
```

Store PCA results and merge with metadata
```{r}
df_pcares <- 
  pca1$x %>% 
  as_tibble(rownames = "Sample") %>% 
  select(Sample,PC1,PC2)
head(df_pcares)

df_pca_res <- 
  df_pcares %>% 
  full_join(df_meta,by = "Sample")

df_pca_res %>% 
  ggplot(aes(x=PC1,y=PC2,color=group)) +
  geom_point() +
  stat_ellipse() +
  theme_bw() +
  theme(aspect.ratio = 1)

df_pca_res %>% 
  ggplot(aes(x=PC1,y=PC2,color=strain_type)) +
  geom_point() +
  stat_ellipse() +
  theme_bw() +
  theme(aspect.ratio = 1)
```
Check outliers
```{r}
outlier_samples <- 
  df_pca_res %>% 
  filter(PC1>5 & PC2<0) %>% 
  pull(Sample)

df_strain_long %>% 
  filter(Sample %in% outlier_samples) %>% 
  pivot_wider(names_from = phenotypes,values_from = value) %>% head()
```

# UMAP
Run UMAP on log-normalized data and store output
```{r}
set.seed(1980)
umap1 <- umap(na.omit(df_pca),n_components = 3)

df_umap_res <-
  umap1 %>%
  as_tibble(rownames = "Sample") %>%
  rename("UMAP1" = "V1") %>%
  rename("UMAP2" = "V2") %>%
  rename("UMAP3" = "V3") %>%
  full_join(df_meta, by = "Sample")
```

Plot UMAP
```{r}
df_umap_res %>% 
  ggplot(aes(x=UMAP1,y=UMAP2,color=group)) +
  geom_point() +
  stat_ellipse() +
  theme_bw() +
  theme(aspect.ratio = 1)

plotly::plot_ly(df_umap_res, x = ~UMAP1, y = ~UMAP2, z = ~UMAP3, color = df_umap_res$group, colors = c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey")) 

df_umap_res %>% 
  ggplot(aes(x=UMAP1,y=UMAP2,color=strain_type)) +
  geom_point() +
  stat_ellipse() +
  theme_void() +
  theme(aspect.ratio = 1) +
  ggtitle("UMAP")
```

# Violin plots

By group
```{r}
df_strain_long %>% 
  mutate(sugar = group) %>% 
  select(-group) %>% 
  ggplot(aes(x=sugar,y=value, fill = sugar)) +
  geom_jitter()

df_strain_long %>% 
  ggplot(aes(x=group,y=value, fill = group)) +
  geom_violin(draw_quantiles = 0.5) +
    scale_fill_manual(values=c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"))+
 facet_grid(vars(phenotypes), scales = "free_y")+
  theme_light()+
  theme(
    text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 20, color = "black", face = "bold"),
    strip.text.y = element_text(size = 20, color = "black", face = "bold")
  )
#ggsave("groups_phenotype_distribution.svg", width = 500, height = 300, unit = "mm", dpi = 300, path="~/plots")

#Evaluate if the groups are statistically different
cm <- compare_means(value ~ group,  data = df_strain_long, ref.group = "hexose",
              method = "wilcox.test", group.by = "phenotypes")
head(cm)

```

By condition
```{r,fig.width=30}
df_strain_long %>% 
  mutate(condition = factor(condition,levels = c("Glu_20",
                                                 "Glu_30",
                                                 "Glu_65",
                                                 "Xyl_16",
                                                 "Xyl_36",
                                                 "Gal_2",
                                                 "Gal_4.5",
                                                 "Ara_2",
                                                 "Ara_4",
                                                 "Man_10",
                                                 "Man_16",
                                                 "EtOH_45",
                                                 "EtOH_90",
                                                 "NaCl_25",
                                                 "NaCl_80",
                                                 "FrAcid_1",
                                                 "FrAcid_3.5",
                                                 "AcAcid_4.5",
                                                 "AcAcid_6",
                                                 "LacAcid_2",
                                                 "LacAcid_7",
                                                 "LevAcid_2.5",
                                                 "LevAcid_5",
                                                 "HMF_0.5",
                                                 "HMF_6",
                                                 "Furf_1",
                                                 "Furf_3",
                                                 "Van_0.5",
                                                 "Van_2"),ordered = T)) %>% 
  ggplot(aes(x=condition,y=value, fill = condition)) +
  geom_violin() +
  facet_wrap(vars(phenotypes),ncol = 1,scales = "free_y") +
  theme_bw() +
  theme(aspect.ratio = 0.2,
        axis.text.x = element_text(angle = 45,hjust = 1)) +
  ggtitle("All phenotypes",subtitle = "Grouped by strain")+
  stat_compare_means(ref.group = "Glu_20",label = "p.signif",hide.ns = T)

```

# Performance correlations  
```{r}
#run correlations for all the combinations of phenotypes
ggpairs(df_strain, columns = c(7:11),ggplot2::aes(colour=group), palette())

#visualize specific correlations with only two phenotypes per time
#ggscatter by phenotype
ggscatter(df_strain, x = 'umax', y = 'CDW', color = 'group', add = "reg.line", conf.int = FALSE, cor.method = "spearman", size=0.5,  palette = c("#00AFBB","#3581d8","#f2c85b", "#fba465", "#FC4E07", "grey"))+
  stat_cor(aes(color = group)) +
  yscale("log10", .format = TRUE)+
  # xscale("log10", .format = TRUE)+
  stat_cor(method = "spearman")+
  theme_light()+
  theme(
    text = element_text(size = 7),
    axis.text.x = element_text(size = 7, angle=90, vjust = 0.5,hjust = 1),
    axis.text.y = element_text(size = 7),
    axis.title = element_text(size = 7, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    strip.text.x = element_text(size = 7, color = "black", face = "bold"),
    strip.text.y = element_text(size = 7, color = "black", face = "bold"))

#ggsave("umax_CDW_corr.svg", width = 150, height = 110, unit = "mm", dpi = 300, path="~/plots")

```





