---
title: "strain_phenotype_calc"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---
Chalmers University of Technology
Gothenburg, Sweden
2023

This script is used to calculate phenotypic variables from growth curves and sugar/ethanol concentrations in high throughput screenings
In the first part the growth variables (maximum specific growth rate, lag phase) are calculated from an excel file exported from the growth profiler in GV units
Each plate analyzed correspond to a specific strain. In the plate various growth conditions are present (plate_layout).

#Loading libraries

```{r}
library(tidyverse)
library(lattice)
library(deSolve)
library(growthrates)
library(ggplot2)
library(ggExtra)
library(naniar)
library(ggpubr)
library(statmod)
```

#Loading data
```{r}
#summary contains information about how the growth_profiler plate is structured
summary <- readxl::read_excel("~/raw_data/plate_legend.xlsx")
#data contains the green values "GV" measured in the growth profiler. Each column corresponds to a well, except the 1st column which is the time in minutes.
#the time column is converted in hours, the last 9 column are removed because they are empty and the plate is reorganized to have only 3 columns (well, values, time(h))
data <- readxl::read_excel("~/raw_data/strain-name.xlsx", sheet = "GV") %>%
  mutate(timeh = `time`/60) %>%
  select(-`time`, -`H4`, -`H5`,-`H6`,-`H7`, -`H8`, -`H9`,-`H10`,-`H11`, -`H12`) %>%
  gather("well", "value", -timeh) 
#the summary of plate and the log of the growth values are merged in the same table, the mean and the standard deviation of the replicates are calculated
strain <- inner_join(data, summary, by="well") %>%
  group_by(condition, timeh) %>%
  mutate(log_value=log(value)) %>%
  mutate(mean=mean(log_value)) %>%
  mutate(stdev=sd(log_value)) %>%
  filter(replicate == 1) 
```

#Statistical test
growth curves from different media are compared with the function "CompareGrowthCurves"
```{r}
stats <- compareGrowthCurves(strain$condition,as.matrix(strain$mean)) %>%
  filter(Group2 == "Glu_20" | Group1 =="Glu_20")
#swapping columns because the output of comparegrowthcurves does not give the right output to plot the data
tmp <- stats[1:12,1]
stats[1:12,1] <- stats[1:12,2]
stats[1:12,2] <- tmp
#plotting P-values
ggboxplot(stats, x = "Group1", y = "P.Value", colour= "Group2", palette = "jco",
          add = "jitter")
#extracting conditions that are significantly different from Glu_20
stats_results <- stats %>%
  filter(P.Value <= 0.05)
print(stats_results$Group2)
```

#Plotting growth curves 
Three plotting options are available to visualize 1)standard deviation among triplicates; 2)all the replicates independently; 3) groups of conditions plotted together

```{r}
#create a table with all replicates to plot in log scale 
strain_replicates <- right_join(data, summary, by="well") %>%
  mutate(replicate= as.character(replicate)) %>%
  mutate(log_value=log(value))

#plot the data visualizing the standard deviation 
palette <- c("#66c2a5", "#fc8d62", "#8da0cb", "#fb8072", "#a6d854", "#ffd92f")

ggplot(data = strain) +
  geom_ribbon(aes(x= timeh, ymin = log_value-stdev, ymax = log_value+stdev, fill= group), alpha= 0.5)+
  geom_line(aes(y=log_value, x=timeh, colour=group)) +
  scale_fill_manual(values = palette) +
  scale_colour_manual(values = palette) +
  labs(title = "LBCM1106",
       subtitle = "growth curves",
       y = "log_GV", x = "time(h)") + 
  facet_wrap(~condition)  +
  theme(legend.title = element_blank(), 
              text = element_text(size = 20),
              panel.background = element_blank(),
              panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                              colour = "grey"), 
              panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                              colour = "grey"))

#ggsave("growth_curves_strain-name.png", width = 10, height = 10, path = "~/plots")
#plot growth curves to visualize replicates
ggplot() +
  geom_point(data = strain_replicates, size = 0.5, aes(y=log_value, x=timeh, colour=group, shape=replicate)) +
  scale_colour_manual(values = palette) +
  labs(title = "LBCM1106",
       subtitle = "growth curves + replicates",
       y = "log_GV", x = "time(h)") + 
  facet_wrap(~condition)  +
  theme(legend.title = element_blank(), 
        text = element_text(size = 20),
        panel.background = element_blank(),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "grey"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "grey"))

#ggsave("growth_curves_replicates_strain-name.png", width = 10, height = 10, path = "~/plots")
#plotting to visualize the groups 
library(RColorBrewer)
nb.cols <- 29
mycolors <- colorRampPalette(brewer.pal(12, "Paired"))(nb.cols)

ggplot(data = strain) +
    geom_ribbon(aes(x= timeh, ymin = mean-stdev, ymax = mean+stdev, fill= condition), alpha= 0.5)+
    geom_line(aes(y=mean, x=timeh, colour=condition)) +
    scale_fill_manual(values = mycolors) +
    scale_colour_manual(values = mycolors)+
    labs(title = "LBCM1106",
         subtitle = "growth curves",
         y = "logGV", x = "time(h)") +
      facet_wrap(~group)  +
  theme(legend.title = element_blank(), 
        text = element_text(size = 20),
        panel.background = element_blank(),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "grey"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "grey"))

#ggsave("growth_curves_groups_strain-name.png", width = 10, height = 10, path = "~/plots")
```

#Calculation of maximum specific growth rate and lag phase 

```{r}
#the function all_splines is used to calculate the µmax 
many_spline_fits <- all_splines(value ~ timeh | group + condition + replicate,
                                data = strain_replicates, spar = 0.7)
par(mfrow = c(6,6))
par(mar=c(1,1,1,1))
plot(many_spline_fits)
#data frame is created with the results coming from all_splines function
results <- results(many_spline_fits)

#calculate lambda (lag phase): the coordinates of the inflection point have been extracted and used to calculate the tangent and subsequently the x value correspondent to the lag phase
i=0
inf_points <- NULL
for (i in 1:87){
  tmp <- many_spline_fits@fits[[i]]@xy 
  inf_points <- rbind(inf_points, tmp)
}
inf_points <- as.data.frame(inf_points)
lambda_mu <- bind_cols(results, inf_points) %>% 
  dplyr::rename(x = V1, y = V2, µmax=mumax) %>%
  dplyr::mutate(lag=((log10(y0)-log10(y))/µmax)+x) %>%
  dplyr::mutate(replicate= as.character(replicate))

#all_splines function will approximate each curve. In the wells with no growth the values reported 
#are wrongly estimated but R2 values are low therefore can be excluded 
lambda_mu$µmax[lambda_mu$r2< 0.985] <- 0
lambda_mu$lag[lambda_mu$r2<0.985] <-0
lambda_mu <- lambda_mu %>%
  select(-`y0`, -`x`, -`y`)

#saving the dataset for further analysis
# write_csv(lambda_mu, path = "~/processed_data/strain-name_growth.csv")
```

#Calculation of yields and cell dry weight

The following part of the script is meant to calculate biomass_yield, ethanol yield and CDW from an excel file containing the summary from OD measurements and enzymatic essays

load data
```{r}
strain_summary <- readxl::read_excel("~/raw_data/strain_summary.xlsx")
summary_dilutions <- readxl::read_excel("~/plate_layout/summary_dilutions.xlsx")
sugar_initial_concentration <- readxl::read_excel("~/plate_layout/sugar_initial_concentration.xlsx")
```
calculation of maximum specific grwoth rate
```{r}
#calculate CDW with calibration curves previously made
CDW <- select(strain_summary, well, condition, replicate, group, final_OD, CDW_OD) %>%
  mutate(CDW=final_OD*CDW_OD[1]) %>%
  slice(1:87)
#final Cell dry weight in g/L with the slope value from the calibration curve
```
glucose concentrations 
```{r}
#calculate glucose concentrations, postion 89 is the standard for the sugar essays
Glu_conc <- select(strain_summary, well, condition, replicate, group, A1_glu, A0_glu)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`man`, -`gal`, -`xyl`, -`ara`) %>%
  mutate(final_glu_conc=((A1_glu-A0_glu)/(A1_glu[89]-A0_glu[89]))*glu*0.4) %>%
  slice(1:87) #to remove the blank and the standards 
```
other sugars concentrations
```{r}
#calculte other sugars concentrations 
Man_conc <- select(strain_summary, well, condition, replicate, group, A1_man, A0_man)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`xyl`, -`ara`) %>%
  mutate(final_man_conc=((A1_man-A0_man)/(A1_man[32]-A0_man[32]))*man*0.1) %>%
  slice(25:30) 
Xyl_conc <- select(strain_summary, well, condition, replicate, group, A1_xyl, A0_xyl)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`man`, -`ara`) %>%
  mutate(final_xyl_conc=((A1_xyl-A0_xyl)/(A1_xyl[14]-A0_xyl[14]))*xyl*0.25) %>%
  slice(7:12)
Gal_conc <- select(strain_summary, well, condition, replicate, group, A1_gal, A0_gal)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`man`, -`xyl`, -`ara`) %>%
  mutate(final_gal_conc=((A1_gal-A0_gal)/(A1_gal[20]-A0_gal[20]))*gal*0.4) %>%
  slice(13:18)
Ara_conc <- select(strain_summary, well, condition, replicate, group, A1_ara, A0_ara)%>%
  inner_join(summary_dilutions, by = c("well", "condition", "replicate", "group")) %>%
  select(-`glu`, -`gal`, -`xyl`, -`man`) %>%
  mutate(final_ara_conc=((A1_ara-A0_ara)/(A1_ara[26]-A0_ara[26]))*ara*0.4) %>%
  slice(19:24)
```
calculate biomass and ethanol yield: g-biomass formed or g-ethanol formed / g-substrate consumed 
```{r}
#create a dataset by marging the different tables calculated before
yields <- select(strain_summary, well, condition, replicate, group, etoh_conc) %>%
  right_join(CDW, by=c("well", "condition", "replicate", "group")) %>%
  select(-`final_OD`, -`CDW_OD`) %>%
  right_join(Glu_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_glu`, -`A0_glu`, -`glu`)%>%
  left_join(Man_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_man`, -`A0_man`, -`man`) %>%
  left_join(Xyl_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_xyl`, -`A0_xyl`, -`xyl`) %>%
  left_join(Gal_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_gal`, -`A0_gal`, -`gal`) %>%
  left_join(Ara_conc, by = c("well", "condition", "replicate", "group")) %>%
  select(-`A1_ara`, -`A0_ara`, -`ara`) %>%
  replace_with_na_all(condition = ~.x <= 0) %>% #replacing negative value with 0 (in the test it means there is no more sugar to measure)
  replace_na(list(final_glu_conc=0, final_man_conc = 0, final_xyl_conc = 0, final_gal_conc = 0, final_ara_conc = 0 )) %>% #replacing NA values with 0 otherwise it cannot calculate the sum
  left_join(sugar_initial_concentration, by=c("well", "condition", "replicate", "group")) %>%
  mutate(final_sugars_conc=final_man_conc+final_xyl_conc+final_gal_conc+final_ara_conc) %>%
  select(-`final_man_conc`,-`final_gal_conc`,-`final_xyl_conc`,-`final_ara_conc`) %>%
  mutate(consumed_glucose=glu_conc-final_glu_conc) %>%
  select(-`glu_conc`,-`final_glu_conc`) %>%
  mutate(consumed_sugarx=other_sugar_conc-final_sugars_conc) %>%
  select(-`other_sugar_conc`,-`final_sugars_conc`) %>%  #calculating how much sugar has been consumed
  replace_with_na_all(condition = ~.x <= 0)%>%  #replacing negative values with 0 
  replace_na(list(consumed_glucose=0, consumed_sugarx = 0, etoh_conc=0)) %>%
  mutate(Yx=CDW/(consumed_glucose+consumed_sugarx)) %>% 
  mutate(Yp=etoh_conc/(consumed_glucose+consumed_sugarx)) %>%
  select(-`etoh_conc`, -`consumed_glucose`,-`consumed_sugarx`)
```

Trimming dataset
```{r}
yields$Yx[yields$Yx == Inf] <-0
yields$Yp[yields$Yp == Inf] <-0
yields$Yp[yields$condition == "EtOH_45"] <-0
yields$Yp[yields$condition == "EtOH_90"] <-0
```

The following part of the script analyses the second cultivation maximum specific growth rate and creates a unique dataset with all the calculated phenotypes. 
```{r}
#analysing the growth of the second growth as above 
data2 <- readxl::read_excel("~/raw_data/strain_2nd growth.xlsx", sheet = "GV") %>%
  mutate(timeh = `time`/60) %>%
  select(-`time`, -`H4`, -`H5`,-`H6`,-`H7`, -`H8`, -`H9`,-`H10`,-`H11`, -`H12`) %>%
  gather("well", "value", -timeh) 

```

merging the growth gata with the metadata
```{r}
#the summary of plate and the growth values are merged in the same table, the mean and the standard deviation of the replicates are calculated
strain2 <- right_join(data2, summary, by="well") %>%
  group_by(condition, timeh) %>%
  mutate(log_value=log(value)) %>%
  mutate(mean=mean(log_value)) %>%
  mutate(stdev=sd(log_value)) 
```
Plotting the 2nd cultivation growth curves 
```{r}
#create a table with all replicates to plot in log scale 
strain_replicates2 <- right_join(data2, summary, by="well") %>%
  mutate(replicate= as.character(replicate)) %>%
  mutate(log_value=log(value))

#plot the data visualizing the standard deviation 
ggplot() +
  geom_point(data = strain_replicates2, size = 0.5, aes(y=log_value, x=timeh, colour=group, shape=replicate)) +
  scale_colour_manual(values = palette) +
  labs(title = "LBCM1106",
       subtitle = "growth curves + replicates",
       y = "log_GV", x = "time(h)") + 
  facet_wrap(~condition)  +
  theme(legend.title = element_blank(), 
        text = element_text(size = 20),
        panel.background = element_blank(),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = "grey"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                        colour = "grey"))
#ggsave("strain_2nd_growth.png", width = 10, height = 10, path = "~/plots")

```

Calculate umax and lag phase in teh second cultivation 
```{r}
#fit the curves and calculate µmax, through the graphs it is possible to visualize the inflection point and the smoothing splines
many_spline_fits_2 <- all_splines(value ~ timeh | group + condition + replicate,
                                  data = strain_replicates2, spar = 0.5)
par(mfrow = c(6,6))
par(mar=c(1,1,1,1))
plot(many_spline_fits_2)
lambda_mu_2 <- results(many_spline_fits_2)%>%
  rename(µmax=mumax)

#all_splines function will approximate each curve. In the wells with no growth the values reported 
#are therefore wrong but R2 values are very low therefore can be excluded 
lambda_mu_2$µmax[lambda_mu_2$r2< 0.985] <- 0
lambda_mu_2$replicate <- as.character(lambda_mu_2$replicate)
```

Merging first growth, yields and second growth
```{r}
growth1 <- lambda_mu
growth2 <- lambda_mu_2 %>%
  rename(µmax2 = µmax) %>%
  inner_join(growth1, by=c("group", "condition", "replicate")) %>%
    select(-`y0`,-`r2.x`,-`r2.y`)
#putting together the growth variables and the production variables
yields$replicate <- as.character(yields$replicate)

strain_data_exp <- yields %>%
  inner_join(growth2, by=c("condition", "replicate","group")) %>%
  # select(-`well`, -`r2`) %>%
  mutate(strain = "strain_name") %>%
  mutate(strain_type ="strain_type") 

#correcting phenotypes when growth is 0
strain_data_exp$CDW[strain_data_exp$µmax == 0] <-0
strain_data_exp$Yx[strain_data_exp$µmax == 0] <-0
strain_data_exp$Yp[strain_data_exp$µmax == 0] <-0
#saving phenotypes for each strain
#write_csv(LBCM1106_data_exp, path = "~/processed_data/strain-name_phenotype.csv")

```


